\chapter{Acknowledgements}


First I would like to express my sincere gratitude to my thesis supervisor Professor Ørnulf Borgan of the Faculty of Mathematics at the University of Oslo, for introducing me to an interesting topic and sharing his knowledge and experience during this past year. The door to Prof. Borgan's office was always open whenever I ran into a trouble spot or had a question about my research or writing.\\

The years at the University of Oslo would have been significantly less interesting without my fellow students. Thank you to Amine, Amanda, Maria and Siren. I appreciate all the insightful discussions and the memorable times we had together. I would also like to thank Prof. Frank Proske for the encouragement and support.
I wish to express my thanks to Kari Mørk for the encouragement and the flexibility that was granted to me at Finance Norway during this time.\\

I must express my very profound gratitude to my family, to the Mehta's, Harriet and Jacob Natvig, Arlette and  Brian for providing me with unfailing support and continuous encouragement throughout my years of study and through the process of researching and writing this thesis.I would like to offer special thanks to Birma for her unceasing support and love.\\

I dedicate this thesis in memory of my dear parents Esther and Jonas Fomete, you are gone but made this journey possible. Lastly, and most of all, I would like to thank GOD, " For from him and through him and to him are all things. To him be glory forever. Amen." Rom 11:36 


\rightline{Blindern, October 24, 2019}  
 \rightline{\textit{Joël Fomete Tankmo}} 

 
%I must express my very profound gratitude to my family, especially to the Mehta's and to Harriet and Jacob Natvig for providing me with unfailing support and continuous encouragement throughout my years of study and through the process of researching and writing this thesis.I would like to offer special thanks to Birma for her unceasing support and love.\\